\Large The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events

### 2023-10-11 paper/main.tex

% tables
\newcommand{\stimDescription}{S1}
\newcommand{\regExpTable}{S2}
\newcommand{\posTags}{S3}
\newcommand{\pastKeys}{S4}
\newcommand{\futureKeys}{S5}
\newcommand{\metaAnalysisDatasets}{S6}

% figures
\newcommand{\segments}{S1}
\newcommand{\events}{S2}
\newcommand{\references}{S3}
\newcommand{\MethodsReplExp}{S4}
\newcommand{\targetAsymmetries}{S5}
\newcommand{\hitRates}{S6}
\newcommand{\characterRefs}{S7}
\newcommand{\refAdjacent}{S8}
\newcommand{\referringReferenced}{S9}

@@

\begin{abstract} {\footnotesize{ How much can we infer about the past and future, given our knowledge of the present? Unlike temporally symmetric inferences about simple sequences, inferences about our own lives are asymmetric: we are better able to infer the past than the future, since we remember our past but not our future (i.e., the psychological arrow of time). What happens when both the past and future are unobserved, as when we make inferences about \textit{other} people's lives? We had participants in two experiments view segments of two character-driven television dramas. They wrote out what would happen just before or after each just-watched segment. Participants were better at inferring past (versus future) events. This asymmetry was driven by participants’ reliance on characters’ conversational references in the narrative, which tended to favor the past. We also carried out a meta analysis to estimate the prevalence of these asymmetries in hundreds of millions of dialogues from television shows, popular movies, novels, and written and spoken natural conversations. We found that, on average, references to the past are roughly 1.5--2 times more prevalent in human conversations than references to the future. Our work reveals a temporal asymmetry in how observations of other people’s behaviors can inform us about the past and future.

@@

\section*{Introduction}

What we experience in the current moment tells us about \textit{now}-- but what does it tell us about the past or future? And does the current moment tell us, as human observers, \textit{more} about the past or about the future? One way of examining these questions is to consider highly simplified scenarios that are artificially constructed in the laboratory \citep[e.g., ][]{MaheEtal22}. At one extreme, for deterministic sequences with \textit{known} rules, knowing the current state provides the observer with sufficient information to exactly reconstruct the entire past and future history of the stimulus. At another extreme, for purely random sequences, observing the current state provides no information about the past \textit{or} future.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{intro1}

  \caption{\textbf{Retrodiction, retrospection, and prediction.} In one's own life, one may draw on memory to retrospect (i.e., review or re-evaluate) the past or predict the future. This process is time-asymmetric, since our own past is (typically) observed whereas our future is not. When we make inferences about \textit{other} people's lives, however, we often have uncertainty about both their past and future, since we may have observed neither. We may \textit{retrodict} the unobserved past and predict the unobserved future of other people's lives.}

  \label{fig:intro1}
\end{figure}

Although the psychological arrow of time implies that we should be better able to infer our past than our future, how generally does this temporal asymmetry hold? And does the asymmetry hold only for our own experiences (due to our memories), or is the asymmetry a general property of any real-life event sequence? In real-world situations (and narratives) where we are \textit{equally} ignorant of the past and future, as for \textit{other} people's lives where we lack memories of the relevant past, are our inferences about the past and future symmetric or asymmetric? For example, imagine that you are meeting a stranger for the first time. At the moment of your meeting, you lack both memories of their past and knowledge about what they might do in the future. After your first encounter with the stranger, would you be able to more accurately or easily form inferences about what had happened in their past (\textit{retrodiction}) or what will happen in their future (\textit{prediction}; Fig.~\ref{fig:intro1})? Or suppose you started watching a movie partway through. Again, you would enter the moment of watching without memories of prior parts of the movie. Given your observations in the present, would your guesses about what had happened before you started watching be more (or less) accurate than your guesses about what will happen next? In general, when the past and future are \textit{both} unobserved, are we better at inferring the past or the future in real-world settings? Narrative stimuli, such as stories and movies, can provide a useful testbed for exploring several of these questions.

@@

One key difference between simple artificial sequences and more naturalistic (real or narrative) sequences is that naturalistic sequences often incorporate other people. Despite the past and future being equally unknown to \textit{the observer} prior to the current moment, other people, and realistic characters in narratives, have their own psychological arrows of time. Specifically, they have memories of their own pasts. Other people's asymmetric knowledge about their \textit{own} pasts and futures might affect their behaviors (e.g., conversations). In turn, this might provide time-asymmetric clues that favor the past~\citep[e.g., other people might talk more about their own pasts than their futures; ][]{DemiEtal18}. If observers leverage these clues from other people's asymmetric knowledge, then observers should also be better at inferring the past (versus the future) of other people's lives. Alternatively, inferences about other people's lives may be more like inferences about artificial statistical sequences~\citep[e.g., perhaps solely relying on statistical regularities like event schemas, scripts, or situation models;][]{RadvCope06, ZwaaRadv98, BoweEtal79, RangRitc12, BaldEtal18}. If so, then the accuracy of inferences about the past and the future of others' lives should be approximately equal. We note that the aforementioned authors make no specific claims about temporal symmetries or asymmetries. Rather, we claim that statistical regularities might \textit{imply} symmetry (e.g., if you are on step $n$ of an unfolding schema, this suggests you have just completed step $n - 1$ and that you are likely to next encounter step $n + 1$).

@@

Participants in our main experiment ($n = 36$) watched segments from two storylines, drawn from the CBS television show \textit{Why Women Kill}. Each storyline comprised 11 segments (mean duration: 2.05~min; range: 0.97--3.87~min, Table~\stimDescription). We asked participants to use free-form (typed) text responses to retrodict what had happened prior to a just-watched segment, predict what would happen next, or recall what they had just watched (Fig.~\ref{fig:method}, \textit{Task design}). We referred to the to-be-retrodicted, to-be-predicted, or to-be-recalled segment as the \textit{target segment} for each response. We systematically varied whether participants watched the segments in forward or reverse chronological order, and how many segments they had seen prior to making a response (see \textit{Methods}).

@@

  \caption{\textbf{Task overview.} Participants in our main experiment watched segments of two storylines from the television series \textit{Why Women Kill}. They made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment. Across four counterbalanced sequences, we systematically varied whether participants watched the segments in forward or reverse chronological order, whether (or not) responses were cued using the main characters in the target segment, and which other segments participants had watched prior to making a response. For each segment, we collected several retrodiction, prediction, and/or recall responses across different experimental conditions. Experiment time is denoted along the vertical axis, storyline segments are indicated along the horizontal axis, and the colors denote experimental tasks (conditions). For an analogous depiction of our replication experiment's design, see Fig.~\MethodsReplExp.}

@@

We asked participants in our main experiment to generate four types of responses after watching each video segment: uncued responses, character-cued responses, updated responses, and recalls (Fig.~\ref{fig:method}, \textit{Data overview}). To generate \textit{uncued} responses, we asked participants to either retrodict (uncued retrodiction; \textit{u-R}) what happened shortly before or predict (uncued prediction; \textit{u-P}) what happened shortly after the just-watched segment. To generate \textit{character-cued} responses, we asked participants to retrodict (character-cued retrodiction; \textit{c-R}) or predict (character-cued prediction; \textit{c-P}) what came before or after the just-watched segment, but we provided additional information to the participant about which character(s) would be present in the target (to-be-retrodicted or to-be-predicted) segment. We hypothesized that character-cued responses should be more accurate than uncued responses, to the extent that participants incorporate the character information we provided to them into their retrodictions and predictions. To generate updated responses, we asked participants to watch an additional segment that came just prior to or just after the target segment, and then to update their retrodiction (\textit{c-RP}) or prediction (\textit{c-PR}) about the target segment. Results on updated responses are not reported in this paper. Finally, we also asked participants to \textit{recall} what happened in the just-watched segment. We labeled these responses according to which other segments participants had watched prior to the just-watched target. Retrodiction-matched recall (\textit{re(R)}) responses were made during the retrodiction sequences (B1 and B2; Fig.~\ref{fig:method}), whereas prediction-matched recall (\textit{re(P)}) responses were made during the prediction sequences (A1 and A2; Fig.~\ref{fig:method}). Whereas retrodiction and prediction responses reflect what participants \textit{estimate} they would remember after watching the (inferred) target segment, recall responses provide a benchmark for comparison by measuring what they \textit{actually} remember about the target segment. Our replication experiment (Fig.~\MethodsReplExp) used a similar design, but did not have participants generate recall, re(R), or re(P) responses.

For each retrodiction and prediction, participants were asked to generate at least one, and not more than three, responses that constituted ``the sorts of things [the participant would] expect to have remembered if [they] had watched the [target] segment.'' They were asked to generate multiple responses only if those additional responses were (in their judgement) of equal likelihood to occur. On average, participants in our main experiment generated 1.08 responses per prompt; therefore we chose to consider only participants' first (``most probable'' or ``most important'') responses to each prompt. We also discarded a small number ($n = 20$) of character-cued responses that did not contain references to all cued characters, along with one additional response due to the participant's misunderstanding of the task instructions during that trial. We carried out our analyses on the remaining 2084 retrodiction, prediction, and recall responses. (Our replication experiment analyses were carried out on XXX responses.)

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.80\textwidth]{results1}

  \caption{\textbf{Retrodiction, prediction, and recall performance by
  experimental condition in our main experiment.} \textbf{A. Methods schematic.} For each retrodiction, prediction, and recall response, we calculated the hit rate for events in the target segment, the response precision (see \textit{Methods}), and the response convergence across participants (see \textit{Methods}). \textbf{B. Target event hit rate.} Mean proportions of target events that were contained in participants' responses, for each response type, averaged across target segments. \textbf{C. Response precision.} Mean precisions of participants' responses, for each response type, averaged across target segments. The horizontal lines denote the mean pairwise semantic similarities (see \textit{Methods}) across recall responses (re(R): orange; re(P): blue). \textbf{D. Response convergence.} Mean (across-participant) convergence of participants' responses, for each response type, averaged across target segments. All panels: error bars denote bootstrapped 95\% confidence intervals. Asterisks indicate significance in the (generalized) linear mixed models: * denotes $p < 0.05$ and ** denotes $p < 0.01$. See Figure~\targetAsymmetries~for analogous results from our replication experiment.}

@@

We used two general approaches to assess the quality of participants' responses (see \textit{Methods}, Figs.~\ref{fig:result1}A). One approach entailed manually annotating events in the video and counting the number of matched events in participants' responses. We identified a total of 117 unique events reflected across the 22 video segments in our main experiment (range: 3--9 per segment; see \textit{Methods}, Table \stimDescription). We assigned one ``point'' to each of these video events. We also identified 23 additional events in participants' responses that were either summaries of several events or that were partial matches to the manually identified video events. We assigned 0.5 point to each of these additional events. This point system enabled us to compute the numbers and proportions (\textit{hit rates}) of correctly retrodicted, predicted, and recalled events contained in each response. Our second approach entailed using a natural language processing model~\citep{CerEtal18} to embed annotations and responses in a 512-dimensional feature space. This approach was designed to capture conceptual overlap between responses that were not necessarily tied to specific events. To quantify this conceptual overlap, we computed the similarities between the embeddings of different sets of responses. Following \cite{HeusEtal21}, we defined the \textit{precision} of each participants' retrodictions or predictions about a target segment as the median cosine similarities between the embeddings of (a) the participant's retrodiction or prediction response for the target segment and (b) each \textit{other} participant's recalls of the same segment. In other words, precision is designed to measure the extent to which retrodictions and predictions captured the conceptual content that (other) participants remembered. We also developed a related measure, which we call \textit{convergence}, to characterize response similarities across participants. In particular, we defined convergence as the mean cosine similarity between the embeddings of a participant's responses to a target segment and all other participants' responses (of the same type) to the same segment. We analyzed the data using generalized linear mixed models, with participant and stimulus (e.g., target segment) identities as crossed random effects (see \textit{Methods}).

First we sought to validate a main effect of response type (i.e., uncued responses, character-cued responses, and recalls), irrespective of the temporal direction (retrodiction versus prediction). Across these three types of responses, participants have access to increasing amounts of information about the target segment. Therefore, across these response types, we hypothesized that participants' responses should become both more accurate and more convergent across individuals. Consistent with this hypothesis, participants' character-cued retrodictions and predictions were associated with higher target event hit rates than uncued retrodictions and predictions in our main experiment (odds ratio (OR): 2.65, $Z = 4.24$, $p < 0.001$, 95\% confidence interval (CI): 1.69 to 4.16; Fig.~\ref{fig:result1}B). These character-cued responses were also more precise ($b = 0.13$, $t(18.1) = 9.43$, $p < 0.001$, CI: 0.10 to 0.16; Fig.~\ref{fig:result1}C) and convergent across individuals ($b = 0.11$, $t(18.6) = 6.21$, $p < 0.001$, CI: 0.07 to 0.15; Fig.~\ref{fig:result1}D). Relative to character-cued responses, participants' recalls showed higher target event hit rates (OR = 21.83, $Z = 10.61$, $p < 0.001$, CI: 12.35 to 38.59) and were more convergence across individuals ($b = 0.20$, $t(19.4) = 9.10$, $p < 0.001$, CI: 0.16 to 0.25). These results are consistent with the common-sense notion that access to more information about a target segment yields better performance (i.e., higher hit rates, precision, and convergence across individuals). These findings also held for our replication experiment (Fig.~\targetAsymmetries; hit rates of character-cued vs. uncued responses: OR: XXX, $Z = XXX$, $p = XXX$, 95\% confidence interval (CI): XXX to XXX; precisions of character-cued vs. uncued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX; convergence of character-cued vs. uncued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX).

Next we carried out a series of analyses specifically aimed at characterizing temporal direction effects--- i.e, the relative quality of retrodictions versus predictions across different types of responses. We hoped that these analyses might provide insights into our central question about whether inferences about the past and future are equally accurate. Across both uncued and character-cued responses in our main experiment (Fig.~\ref{fig:method}), retrodictions had numerically higher hit rates than predictions (Fig.~\ref{fig:result1}B). However, these differences were only statistically reliable for character-cued responses (uncued responses: OR = 1.17, $Z = 0.35$, $p = 0.73$, CI: 0.47 to 2.92; character-cued responses: OR = 1.93, $Z = 2.15$, $p = 0.03$, CI: 1.06 to 3.52). We observed a similar pattern of results for the precisions of participants' responses (Fig.~\ref{fig:result1}C). Specifically, their responses tended to be numerically more precise for retrodictions versus predictions, but the differences were only statistically reliable for character-cued responses (uncued responses: $b = 0.03$, $t(20.9) = 1.09$, $p = 0.29$, CI: -0.03 to 0.10; character-cued responses: $b = 0.06$, $t(20.8) = 3.01$, $p = 0.007$, CI: 0.02 to 0.11). We also consistently observed numerically higher convergence across participants for retrodictions versus predictions (Fig.~\ref{fig:result1}D), but neither of these differences were statistically reliable (uncued responses: $b = 0.03$, $t(17.9) = 0.75$, $p = 0.46$, CI: -0.05 to 0.11; character-cued responses: $b = 0.04$, $t(17.4) = 1.46$, $p = 0.16$, CI: -0.02 to 0.09). In our replication experiment (Fig.~\targetAsymmetries), participants were numerically better at making \textit{predictions} than retrodictions, but none of these differences were statistically reliable (hit rate for uncued responses: OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX; hit rate for character-cued responses: OR = XXX, $Z = XXX$, $p = XXX$, CI: XXX to XXX; precision for uncued response: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX; precision for character-cued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX; convergence for uncued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX; convergence for character-cued responses: $b = XXX$, $t(XXX) = XXX$, $p = XXX$, CI: XXX to XXX). Taken together, our results across our main and replication experiment suggest that whether participants are better at retrodicting versus predicting the immediate past or future may be somewhat stimulus specific. We also verified that this was not solely a consequence of how participants' memory performance might have been affected by watching different segments (or making different responses to other segments) across conditions by comparing recall responses in the retrodiction-matched recall (\textit{re(R)}) and prediction-matched recall (\textit{re(P)}) conditions. Recall performance in our main experiment was similar in both conditions (target event hit rate: OR = 1.12, $Z = 1.07$, $p = 0.29$, CI: 0.91 to 1.39; convergence: $b = 0.03$, $t(19.3) = 1.89$, $p = 0.07$, CI: 0.00 to 0.07). (We did not collect recall responses in our replication experiment.)

\begin{figure}[tp]
  \centering
  \includegraphics[width=\textwidth]{results2}

>>>>

% tables
\newcommand{\stimDescription}{S1}
\newcommand{\stimDescriptionRep}{S2}
\newcommand{\metaAnalysisDatasets}{S3}
\newcommand{\regExpTable}{S4}
\newcommand{\posTags}{S5}
\newcommand{\pastKeys}{S6}
\newcommand{\futureKeys}{S7}

% figures
\newcommand{\segments}{S1}
\newcommand{\stackedbar}{S4}
\newcommand{\stackedbarRep}{S5}
\newcommand{\refer}{S6}
\newcommand{\referRep}{S7}
\newcommand{\refEffectRep}{S8}

\newcommand{\MethodsReplExp}{S1}
\newcommand{\targetAsymmetries}{S5}
\newcommand{\hitRates}{S6}
\newcommand{\characterRefs}{S7}
\newcommand{\storylines}{S8}
\newcommand{\refAdjacent}{S9}
\newcommand{\refAdjacentCorrected}{S10}
\newcommand{\referringReferenced}{S11}

@@

\begin{abstract} {\footnotesize{ How much can we infer about the past and future, given our knowledge of the present? Unlike temporally symmetric inferences about simple sequences, inferences about our own lives are asymmetric: we are better able to infer the past than the future, since we remember our past but not our future (i.e., the psychological arrow of time). What happens when both the past and future are unobserved, as when we make inferences about \textit{other} people's lives? We had participants in two experiments view segments of two character-driven television dramas. They wrote out what would happen just before or after each just-watched segment. Participants were better at inferring past (versus future) events. This asymmetry was driven by participants’ reliance on characters’ conversational references in the narrative, which tended to favor the past. We also carried out a meta analysis to estimate the prevalence of temporal asymmetries in past versus future conversational references in hundreds of millions of dialogues from television shows, popular movies, novels, and written and spoken natural conversations. We found that, on average, references to the past are 1.45 times more prevalent in human conversations than references to the future. Our work reveals a temporal asymmetry in how observations of other people’s behaviors can inform us about the past and future.

@@

\section*{Introduction}
What we experience in the current moment tells us about \textit{now}-- but what does it tell us about the past or future? And does the current moment tell us, as human observers, \textit{more} about the past or about the future? One way of examining these questions is to consider highly simplified scenarios that are artificially constructed in the laboratory \citep[e.g., ][]{MaheEtal22}. At one extreme, for deterministic sequences with \textit{known} rules, knowing the current state provides the observer with sufficient information to exactly reconstruct the entire past and future history of the stimulus. At another extreme, for purely random sequences, observing the current state provides no information about the past \textit{or} future.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.6\textwidth]{intro}

  \caption{\textbf{Retrodiction, retrospection, and prediction.} In one's own life, one may draw on memory to retrospect (i.e., review or re-evaluate) the past or predict the future. This process is time-asymmetric, since our own past is (typically) observed whereas our future is not. When we make inferences about \textit{strangers'} lives, however, we often have uncertainty about both their past and future, since we may have observed neither. We may \textit{retrodict} the unobserved past and predict the unobserved future of strangers' lives.}

  \label{fig:intro}
\end{figure}

Although the psychological arrow of time implies that we should be better able to infer our past than our future, how generally does this temporal asymmetry hold? And does the asymmetry hold only for our own experiences (due to our memories), or is the asymmetry a general property of any real-life event sequence? In real-world situations (and narratives) where we are \textit{equally} ignorant of the past and future, as for \textit{other} people's lives where we lack memories of the relevant past, are our inferences about the past and future symmetric or asymmetric? For example, imagine that you are meeting a stranger for the first time. At the moment of your meeting, you lack both memories of their past and knowledge about what they might do in the future. After that first encounter with the stranger, would you be able to more accurately or easily form inferences about what had happened in their past (\textit{retrodiction}) or what will happen in their future (\textit{prediction}; Fig.~\ref{fig:intro})? Or suppose you started watching a movie partway through. Again, you would enter the moment of watching without memories of prior parts of the movie. Given your observations in the present, would your guesses about what had happened before you started watching be more (or less) accurate than your guesses about what will happen next? In general, when the past and future are \textit{both} unobserved, are we better at inferring the past or the future in real-world settings? Narrative stimuli, such as stories and movies, can provide a useful testbed for exploring several of these questions.

@@

One key difference between simple artificial sequences and more naturalistic (real or narrative) sequences is that naturalistic sequences often incorporate other people. Despite the past and future being equally unknown to \textit{the observer} prior to the current moment, other people, and realistic characters in narratives, have their own psychological arrows of time. Specifically, they have memories of their own pasts. Other people's asymmetric knowledge about their \textit{own} pasts and futures might affect their behaviors (e.g., conversations). In turn, this might provide time-asymmetric clues that favor the past~\citep[e.g., other people might talk more about their own pasts than their futures; ][]{DemiEtal18}. If observers leverage these clues from other people's asymmetric knowledge, then observers should also be better at inferring the past (versus the future) of other people's lives. Alternatively, inferences about other people's lives may be more like inferences about artificial statistical sequences~\citep[e.g., perhaps solely relying on statistical regularities like event schemas, scripts, or situation models;][]{RadvCope06, ZwaaRadv98, BoweEtal79, RangRitc12, BaldEtal18}. If so, then the accuracy of inferences about the past and the future of others' lives should be approximately equal. We note that the aforementioned authors make no specific claims about temporal symmetries or asymmetries. Rather, we claim that statistical regularities might \textit{imply} symmetry (e.g., if you are on step $n$ of an unfolding schema, this suggests you may have just completed step $n - 1$ and that you may next encounter step $n + 1$).

@@

Participants in our main experiment ($n = 36$) watched segments from two storylines, drawn from the CBS television show \textit{Why Women Kill}. Each storyline comprised 11 segments (mean duration: 2.05~min; range: 0.97--3.87~min, Table~\stimDescription). We asked participants to use free-form (typed) text responses to retrodict what had happened prior to a just-watched segment, predict what would happen next, or recall what they had just watched (Fig.~\ref{fig:method}, \textit{Task design}). We referred to the to-be-retrodicted, to-be-predicted, or to-be-recalled segment as the \textit{target segment} for each response. We systematically varied whether participants watched the segments in forward or reverse chronological order, and how many segments they had seen prior to making a response (see \textit{Task design and procedure}).  We also ran a pre-registered replication study with a similar design, where participants ($n = 37$) watched segments from the Netflix television show \textit{The Chair}, comprising 13 segments (mean duration: 1.97~min; range: 0.58--4.30~min, Table~\stimDescriptionRep).

@@

  \caption{\textbf{Task overview.} Participants in our main experiment watched segments of two storylines from the television series \textit{Why Women Kill}. They made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment. Across four counterbalanced sequences, we systematically varied whether participants watched the segments in forward or reverse chronological order, whether (or not) responses were cued using the main characters in the target segment, and which other segments participants had watched prior to making a response. For each segment, we collected several retrodiction, prediction, and/or recall responses across different experimental conditions. Experiment time is denoted along the vertical axis, storyline segment orders are indicated along the horizontal axis, and the colors denote experimental tasks (conditions). For an analogous depiction of our replication experiment's design, see Fig.~\MethodsReplExp.}

@@

We asked participants in our main experiment to generate four types of responses after watching each video segment: uncued responses, character-cued responses, updated responses, and recalls (Fig.~\ref{fig:method}, \textit{Data overview}). To generate \textit{uncued} responses, we asked participants to either retrodict (uncued retrodiction; \textit{u-R}) what happened shortly before or predict (uncued prediction; \textit{u-P}) what happened shortly after the just-watched segment. To generate \textit{character-cued} responses, we asked participants to retrodict (character-cued retrodiction; \textit{c-R}) or predict (character-cued prediction; \textit{c-P}) what came before or after the just-watched segment, but we provided additional information to the participant about which character(s) would be present in the target (to-be-retrodicted or to-be-predicted) segment. We hypothesized that character-cued responses should be more accurate than uncued responses, to the extent that participants incorporate the character information we provided to them into their retrodictions and predictions. To generate updated responses, we asked participants to watch an additional segment that came just prior to or just after the target segment, and then to update their retrodiction (\textit{c-RP}) or prediction (\textit{c-PR}) about the target segment. Results on updated responses are not reported in this paper. Finally, we also asked participants to \textit{recall} what happened in the just-watched segment. We labeled these responses according to which other segments participants had watched prior to the just-watched target. Retrodiction-matched recall (\textit{re(R)}) responses were made during the retrodiction sequences (B1 and B2; Fig.~\ref{fig:method}), whereas prediction-matched recall (\textit{re(P)}) responses were made during the prediction sequences (A1 and A2; Fig.~\ref{fig:method}). Whereas retrodiction and prediction responses reflect what participants \textit{estimate} they would remember after watching the (inferred) target segment, recall responses provide a benchmark for comparison by measuring what they \textit{actually} remember about the target segment. Our replication experiment (Fig.~\MethodsReplExp) used a similar design, but did not have participants generate recall responses.

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.7\textwidth]{results1}

  \caption{\textbf{Retrodiction, prediction, and recall performance by experimental condition in our main and replication experiments.} \textbf{A. Methods schematic.} For each retrodiction, prediction, and recall response, we calculated the hit rate for events in the target segment (see \textit{Response analyses}), the response precision (see \textit{Text embeddings of participants’ responses}), and the response convergence across participants. \textbf{B. Target event hit rate (main experiment).} Mean proportions of target events that were contained in participants' responses, for each response type, averaged across target segments. \textbf{C. Response precision (main experiment).} Mean precisions of participants' responses, for each response type, averaged across target segments. The horizontal lines denote the mean pairwise semantic similarities (see \textit{Text embeddings of participants’ responses}) across recall responses (re(R): orange; re(P): blue). \textbf{D. Response convergence (main experiment).} Mean (across-participant) convergence of participants' responses, for each response type, averaged across target segments. \textbf{E. Target event hit rate (replication experiment)}. Same format as Panel B. 
 \textbf{F. Response precision (replication experiment)}. Same format as Panel C.  \textbf{G. Response convergence (replication experiment)}. Same format as Panel D. All panels: error bars denote bootstrapped 95\% confidence intervals. Asterisks indicate significance in the (generalized) linear mixed models: * denotes $p < 0.05$ and ** denotes $p < 0.01$.}

@@

We used two general approaches to assess the quality of participants' responses (see \textit{Response analyses, Text embeddings of participants’ responses}, Fig.~\ref{fig:result1}A). One approach entailed manually annotating events in the video and counting the number of matched events in participants' responses. We identified a total of 117 unique events reflected across the 22 video segments in our main experiment (range: 3--9 per segment), and a total of 71 events across the 13 segments in our replication experiment (range 1--16; see \textit{Video annotation}, Tables~\stimDescription,~\stimDescriptionRep). We assigned one ``point'' to each of these video events. We also identified a number of additional events (main experiment: 23; replication experiment: 17) in participants' responses that were either summaries of several events or that were partial matches to the manually identified video events. We assigned 0.5 point to each of these additional events. This point system enabled us to compute the numbers and proportions (\textit{hit rates}) of correctly retrodicted, predicted, and recalled events contained in each response. Our second approach entailed using a natural language processing model~\citep{CerEtal18} to embed annotations and responses in a 512-dimensional feature space. This approach was designed to capture conceptual overlap between responses that were not necessarily tied to specific events. To quantify this conceptual overlap, we computed the similarities between the embeddings of different sets of responses. Following \cite{HeusEtal21}, we defined the \textit{precision} of participants' responses as the median cosine similarity between that response's vector and the embedding vectors for all other participants' recalls of the target segment (main experiment), or the similarity between that response's vector and the embedding vector for an online plot synopsis (obtained via Screen Spy; \url{www.screenspy.com/the-chair-season-1-episode-1}) of the target segment (replication experiment).  In other words, precision is designed to measure the extent to which retrodictions and predictions captured the conceptual content that (other) participants remembered. We also developed a related measure, which we call \textit{convergence}, to characterize response similarities across participants. In particular, we defined convergence as the mean cosine similarity between the embeddings of a participant's responses to a target segment and all other participants' responses (of the same type) to the same segment. We analyzed the data using generalized linear mixed models, with participant and stimulus (e.g., target segment) identities as crossed random effects (see \textit{Statistical analysis}).

First we sought to validate a main effect of response type (i.e., uncued responses, character-cued responses, and recalls), irrespective of the temporal direction (retrodiction versus prediction). Across these three types of responses, participants have access to increasing amounts of information about the target segment. Therefore, across these response types, we hypothesized that participants' responses should become both more accurate and more convergent across individuals. Consistent with this hypothesis, participants' character-cued retrodictions and predictions were associated with higher target event hit rates than uncued retrodictions and predictions in our main experiment (odds ratio (OR): 2.65, $Z = 4.24$, $p < 0.001$, 95\% confidence interval (CI): 1.69 to 4.16; Fig.~\ref{fig:result1}B). These character-cued responses were also more precise ($b = 0.13$, $t(18.1) = 9.43$, $p < 0.001$, CI: 0.10 to 0.16; Fig.~\ref{fig:result1}C) and convergent across individuals ($b = 0.11$, $t(18.6) = 6.21$, $p < 0.001$, CI: 0.07 to 0.15; Fig.~\ref{fig:result1}D). Relative to character-cued responses, participants' recalls showed higher target event hit rates (OR = 21.83, $Z = 10.61$, $p < 0.001$, CI: 12.35 to 38.59) and were more convergence across individuals ($b = 0.20$, $t(19.4) = 9.10$, $p < 0.001$, CI: 0.16 to 0.25). These results are consistent with the common-sense notion that access to more information about a target segment yields better performance (i.e., higher hit rates, precision, and convergence across individuals).  These findings also held for our replication experiment (target event hit rates of character-cued vs. uncued responses: OR: 18.63, $Z = 4.26$, $p < 0.001$, CI: 4.85 to 71.58, Fig.~\ref{fig:result1}E; precisions of character-cued vs. uncued responses: $b = 0.26$, $t(11.70) = 9.87$, $p < 0.001$, CI: 0.20 to 0.31, Fig.~\ref{fig:result1}F; convergence of character-cued vs. uncued responses: $b = 0.25$, $t(11.98) = 8.93$, $p < 0.001$, CI: 0.19 to 0.31, Fig.~\ref{fig:result1}G).

Next we carried out a series of analyses specifically aimed at characterizing temporal direction effects--- i.e, the relative quality of retrodictions versus predictions across different types of responses. We hoped that these analyses might provide insights into our central question about whether inferences about the past and future are equally accurate. Across both uncued and character-cued responses in our main experiment (Fig.~\ref{fig:method}), retrodictions had numerically higher hit rates than predictions (Fig.~\ref{fig:result1}B). However, these differences were only statistically reliable for character-cued responses (uncued responses: OR = 1.17, $Z = 0.35$, $p = 0.73$, CI: 0.47 to 2.92; character-cued responses: OR = 1.93, $Z = 2.15$, $p = 0.03$, CI: 1.06 to 3.52). We observed a similar pattern of results for the precisions of participants' responses (Fig.~\ref{fig:result1}C). Specifically, their responses tended to be numerically more precise for retrodictions versus predictions, but the differences were only statistically reliable for character-cued responses (uncued responses: $b = 0.03$, $t(20.9) = 1.09$, $p = 0.29$, CI: -0.03 to 0.10; character-cued responses: $b = 0.06$, $t(20.8) = 3.01$, $p = 0.007$, CI: 0.02 to 0.11). We also consistently observed numerically higher convergence across participants for retrodictions versus predictions (Fig.~\ref{fig:result1}D), but neither of these differences were statistically reliable (uncued responses: $b = 0.03$, $t(17.9) = 0.75$, $p = 0.46$, CI: -0.05 to 0.11; character-cued responses: $b = 0.04$, $t(17.4) = 1.46$, $p = 0.16$, CI: -0.02 to 0.09).  In our replication experiment, as in our main experiment, most of these differences were not statistically reliable (target event hit rates for uncued responses: OR = 0.11, $Z = -1.92$, $p = 0.05$, CI: 0.01 to 1.04; target event hit rates for character-cued responses: OR = 1.42, $Z = 0.62$, $p = 0.53$, CI: 0.47 to 4.23, Fig.~\ref{fig:result1}E; precision for uncued response: $b = -0.06$, $t(15.86) = -1.85$, $p = 0.08$, CI: -0.12 to 0.01; precision for character-cued responses: $b = -0.04$, $t(25.02) = -2.28$, $p = 0.03$, CI: -0.08 to 0.00, Fig.~\ref{fig:result1}F; convergence for uncued responses: $b = -0.03$, $t(12.15) = -0.55$, $p = 0.59$, CI: -0.13 to 0.07; convergence for character-cued responses: $b = -0.05$, $t(13.68) = -1.78$, $p = 0.10$, CI: -0.11 to 0.01, Fig.~\ref{fig:result1}G).  Taken together, our results from both experiments suggest that participants’ inferences about the immediate past and future do \textit{not} show reliable asymmetries.

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.7\textwidth]{results2}

  \caption{\textbf{Retrodictions and predictions of temporally near and distant events.} \textbf{A. Illustration of annotation approach.} For each uncued retrodiction and prediction response, we calculated the number of (retrodicted or predicted) events as a function of temporal distance from the target segment, or \textit{lag}. Onscreen (explicit) events are tagged using integer-valued lags, whereas offscreen (implicit) events are tagged using half-step lags ($\pm 0.5$, $\pm 1.5$, etc.). \textbf{B. Number of events hit in participants' uncued retrodictions and predictions for each event type (main experiment).} Here we separated events we identified in participants' responses according to whether they occurred in the target segment (lags of $\pm 1$), during the interval between the target segment and the just-watched segment (lags of $\pm 0.5$), at longer temporal distances ($|\mathrm{lag}| > 1$), or were incorrect (unmatched with any past or future events in the narrative). The counts displayed in the panel are averaged across just-watched segments. \textbf{C. Number of events hit as a function of temporal distance (main experiment).} Here the (across-segment) mean numbers of events hit in participants' uncued retrodictions (orange) and predictions (blue) are displayed as a function of temporal distance to the just-watched segment (lag). \textbf{D. Number of events hit in participants' uncued retrodictions and predictions for each event type (replication experiment).} Same format as Panel B.  \textbf{E. Number of events hit as a function of temporal distance (replication experiment).} Same format as Panel C.  Error bars denote bootstrapped 95\% confidence intervals. Colors denote temporal direction (orange: past; blue: future) and distance (darker shading: onscreen events from segments adjacent to the target segment; lighter shading: offscreen events).}
